Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2013-05-12 large_number_squeezing_with_atoms.tex
Major rewrite of text for new conclusions
Appendix unexamined, and I may have gone off the rails at several
points.  I put bold statements near the most concerning, but there may
be others.  I think most of Mattias' Bog/TWA figures should be
condensed into two, so the text assumes this.

Will polish after basic truthiness is checked.

At time $t=0$ the coupling $\Omega$ is switched on and applied until time $t=t_1$, at which point it is turned off, resulting in a portion of the population being transferred into mode $|b\rangle$. As the initial state was a coherent state in both modes, after the population transfer both $|a\rangle$ and $|b\rangle$ are described by coherent states, with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.

@@

After this hold time $\tau_{\mathrm{hold}} = t_2 - t_1 $, the coupling $\Omega$ is switched back on until time $t_3$, with a phase shift $\phi$ compared to its first application. During this last stage, the two modes exchange population and the quadrature variance is converted into number variance, in the same way homodyne measurements are used in quantum optics to convert quadrature squeezing into number squeezing, which can be directly measured. In this interpretation, $\phi$ is the relative phase of the strong local oscillator, which allows specific phase angles of quadrature squeezing to be examined.  This experiment can also be interpreted as a Ramsey interferometer with a final beam splitter phase of $\phi$.

@@

This system can be solved analytically, and expressions for quadrature, absolute and relative number squeezing can be derived. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation}.
If we define
\begin{eqnarray}
s &=& \sin(\Omega (t_3 - t_2)) \label{eqsDef} \\
c &=& \cos(\Omega (t_3 - t_2)) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} / \hbar \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after appling the final coupling field are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \nonumber\\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) (n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B 
                     + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^*) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, e.g. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible. It should also be noted that, although there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences. This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

@@

While the analytic expressions for the number variances are not particularly edifying, it is easy to evaluate and plot them for any given set of parameters. As an example, in Figure \ref{figTwoModeAnalyticExamples} we plot the solution for the normalized number variance in mode $|a\rangle$ for a particular set of nonlinearities, choosing a number of different values of $\phi$. The plots illustrate the basic features of the problem: During the final coupling period, the normalised number variance undergoes cyclic variation with a period $\tau=\pi / \Omega$, and has a series of minima. It is necessary to choose both the correct phase $\phi$ as well as the correct length of time for the coupling to be applied in order to obtain optimum squeezing.

This two-mode analytic model allows arbitrarily good number squeezing, provided there are no limits to the number of particles in the system or to the hold time. Unfortunately, in any realistic system, there are more than two modes present, and multimode effects conspire to limit the level of squeezing that can be achieved. It is to this that we now turn.

@@

\begin{itemize}
  \item Maybe point out it agrees with Oberthaler, or save that for discussion on MM effects?
  \item Point out that number *difference* squeezing can often be considerably higher than absolute number squeezing in a single state
\end{itemize}

@@

An analytic solution to the multimode, higher-dimensional problem is intractable, so we must use a numerical solution.  We use stochastic methods based on phase space representations to simulate the dynamics of the multimode quantum fields.  These methods alleviate the problem of directly simulating elements of the Hilbert space, the size of which increases exponentially with the number of spatial modes \cite{stochasticRefs}.  Stochastic methods achieve this by finding a sufficiently well-behaved quasi-probability representation for the density matrix, which can then be simulated as the average behaviour of a number of low-dimensional samples.  These sample objects are the same size as the equivalent classical field.  In our case, the only tractable phase space representation is the functional Wigner representation, which is well behaved when the quantum field is approximately Gaussian, as it is for coherent and squeezed states.  Stochastic methods based on the functional Wigner representation have been used to model the behaviour or BECs and atom lasers \cite{johnssonET2007,dallET2009,dennisET2010}, as the Wigner representation allows the mapping of the system to a Fokker-Plank equation (unlike the $P$-representation), and does not suffer exponential path-weighting problems that would constrain it to very short simulation intervals (as would the positive-$P$ representation).

The equation of motion for the density matrix is defined by the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}).  This defines the evolution of the functional Wigner distribution of the system, which has a one-to-one correspondence with the density matrix.  The equation of motion for the functional Wigner distribution contains derivatives of third order which we assume to be negligible.  This uncontrolled approximation is called the Truncated Wigner Approximation (TWA), and while it has been used widely on ultracold gases \cite{atomlaserWigner,orOtherUCGWigner}, care must be taken that the simulation remains valid. For our simulations we checked the results of our simulations against both the analytic two-mode solution, as well as checking for typical indications of TWA breakdown such a the appearance of negative densities in lightly populated modes. In all the regimes we investigated the TWA remained valid, apart from the most demanding simulations which took place in 3D with high nonlinearities and long hold times. The difficulty with these 3D simulations was that due to the large number of modes they included, the number of particles per mode could drop to ten or less in the densest regions where the squeezing was generated. Over very long time scales, this low mode occupation began to result in TWA breakdown. As we could characterise the regime in which breakdown began to occur, however, we were able stay away from it and to ensure our simulations remained valid.

Under the TWA, the equation of motion for the functional Wigner distribution is a Fokker-Planck equation, and can therefore be sampled by a set of stochastic equations, as described in \cite{GardinerQuantumNoise}.  These stochastic fields $\phi(\mathbf{x})$ are related to field operator expectation values by  
\begin{equation}
\langle :\hat{\psi}^\dagger(\mathbf{x_1})\cdots\hat{\psi}^\dagger(\mathbf{x_n})\hat{\psi}(\mathbf{y_1})\cdots\hat{\psi}(\mathbf{y_m}):_{\mbox{sym}}\rangle = \mathbb{E}\left[\phi^*(\mathbf{x_1})\cdots\phi^*(\mathbf{x_n})\phi(\mathbf{y_1})\cdots\phi(\mathbf{y_m})\right]
\label{eqExpectationRelations}
\end{equation}
%
where $:\star:_{\mbox{sym}}$ denotes symmetric ordering of the operators, and $\mathbb{E}$ denotes a stochastic average over the variables $\phi(\mathbf{x})$.  While we see below that the evolution of these fields is deterministic, the initial state still requires a random element, so multiple realisations are required.  When approximating these equations on a discrete grid, the magnitude of initial noise, the relationship between stochastic averages and the expectation values, and the equations of motion all become grid-dependent.  This is less concerning as it may at first appear, as the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}) assumes a contact potential between the atoms, and this is not correct below a length scale that can probe the details of the true potentials.  Furthermore, well before that regime is reached, the different elements of grid dependence cancel out to give grid-independent predictions for physical observables.

@@

\textbf{Note the initial state we use, and the equations.}

We integrate equations (\ref{eqStochasticEquations}) numerically in one, two and three spatial dimensions using the numerical package XMDS2 \cite{dennis2013}.  For the one- and two-dimensional simulations, the dimensional reduction is achieved by estimating the mode shape in three dimensions, and using dimensionally reduced values for $U_{ij}$ that match the zero-dimensional reduction given by Eq.~(\ref{eqChiUequivalence}).  An equivalent process is to match the chemical potentials of the initial states.

@@

\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/modal_squeezing_effects_1D.eps}
    \caption{Effects of mode shape on squeezing. The plot shows the normalized number variance in mode $|a \rangle$ as a function of the time the final coupling pulse is applied. Thick red line, analytic two-mode solution; solid blue line, constant density mode; dashed line, Thomas-Fermi mode; dash-dotted line, Gaussian mode. The nonlinearities of the various 1D simulations have been adjusted for their mode shape so that they are equivalent to that of the 0D, two-mode mode nonlinearity. Parameters: $n_a = n_b =1 \times 10^5$, $\chi_{aa}=0.03 \hbar, \chi_{ab}=\chi_{bb}=0, \tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=4.0$. Maximum squeezing is achieved when the atomic density profile is the most uniform.}
    \label{figModalSqueezingEffects1D}
\end{figure}

One way to avoid the requirement to engineer and maintain specific atomic modes is to post-select a mode.  Rather than considering the variance in the total number of atoms in a mode, it may be possible to consider only the statistics of a spatially filtered area of the atomic cloud.  For example, suppose the atoms were in a Gaussian atomic distribution, but we consider a mode that only includes the atoms within $p$ standard deviations of the centre. For low $p$, this mode shape is considerably closer to the constant density case than a standard Gaussian. The equivalence between the single mode case and this mode is given by
\begin{equation}
\chi = U \frac{{\mathrm{Erf}} (\sqrt{2p})^3}{{\mathrm{Erf}}(p)^6} \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}}
\end{equation}
A simulations comparing the best attainable squeezing for this mode compared to an unfiltered Gaussian is shown in Figure~\ref{XXX}.  We can see that the squeezing is higher.  One obvious disadvantage to post-selecting modes is that it  reduce the effective number of atoms.  This is a serious issue for most applications of squeezed sources, such as interferometry, where the signal to noise scales with the square root of the flux.

\section{Using a $\pi$-pulse to improve mode-matching}
Modal mismatch is generated by the different atomic potentials seen by the two internal states.  We propose a straightforward method of minimising this effect: using a $\pi$-pulse to swap the populations of the internal states after half of the evolution. As can be seen by Eqs~(\ref{eqa1},\ref{eqb1}, such a pulse also leaves the quantum statistics of each state unchanged. In the limit where each state stayed in a single mode (although that mode might change), and exactly half of the atoms are in each state, this method would produce identical mode shapes for both modes after the hold time. In practice this will never work perfectly as the mode shapes will change, resulting in non-identical environments for the two modes. Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement. More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.

These points are illustrated in Fig.~\ref{piPulseFig}, which shows the best squeezing obtainable from a specific system with and without the $\pi$-pulse, as well as the extra squeezing that can be obtained by allowing a longer hold time as well as the $\pi$-pulse.
\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/pi_pulse.eps}
    \caption{Effect of a $\pi$-pulse halfway through the hold time. Common parameters: $n_a = n_b =2 \times 10^5$, $\chi_{aa}=0.03\hbar$, $\chi_{ab}=\chi_{bb}=0$. Solid red line: Analytic two-mode solution, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.0$ (10.9dB squeezing); Solid blue line: no $\pi$-pulse, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.0$ (4.9dB squeezing);  Dashed blue line: with $\pi$-pulse, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.05$ (5.8dB squeezing); Dash-dotted blue line: with $\pi$-pulse and the longer hold time $\tau_{\mathrm{hold}}=5\times 10^{-4}$s allowed by the application of the pulse, $\phi=3.87$ (7.2dB squeezing).}
    \label{piPulseFig}
\end{figure}

\section{Squeezing in 2D and 3D}

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=7cm]{figures/dimensional_effects_on_squeezing_1.eps}
    \label{figDimensionalSqueezingEffects:sub1}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=7cm]{figures/dimensional_effects_on_squeezing_2.eps}
    \label{figDimensionalSqueezingEffects:sub2}
  \end{subfigure}
\caption{Effects of dimension on squeezing in two different nonlinear regimes. If two nonlinearities are non-zero, the deleterious effects of higher dimensions on squeezing are exacerbated. Thick red line, analytic two-mode solution; solid blue line, 1D; dashed line, 2D; dash-dotted line, 3D. (a) Parameters: $n_a = n_b =2 \times 10^5$, $\chi_{aa}=\chi_{bb}=0.03\hbar$, $\chi_{ab}=0$, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=6.2$. (b) $n_a = n_b =2 \times 10^5$, $\chi_{aa}=0.03\hbar$, $\chi_{bb}=\chi_{ab}=0$, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.1$.}
  \label{figDimensionalSqueezingEffects}
\end{figure}

The simulations in the previous two sections were one dimensional, and show how having spatially constant density and using a $\pi$-pulse will improve the resultant squeezing.  While these conclusions remain valid for simulations in two or three dimensions, adding each additional dimension results in a degradation of the squeezing.  This is true even when the potential (and the resulting atomic density) is constant across the bulk of the trap.  Furthermore, the expected squeezing goes down as the volume of the trap increases.  This can be seen in Fig.~\ref{figDimensionalSqueezingEffects}, where we show the squeezing generated in 1D, 2D and 3D BECs for two different parameter regimes. In both cases we take the BEC to be trapped in a box potential, which in the Thomas-Fermi limit leads to a constant density profile across the BEC. In Figure~\ref{figDimensionalSqueezingEffects}a we consider the case where $U_{aa}=U_{bb}\neq 0$, $U_{ab}=0$, while in Figure~\ref{figDimensionalSqueezingEffects}b we consider the case $U_{aa}\neq 0$, $U_{ab}=U_{bb}=0$. In both cases it is clear that number squeezing is best in lower dimensions, and becomes degraded or non-existent as we move to a three dimensional system; the effect is worse if more than one nonlinearity is non-zero.

>>>>

At time $t=0$, the coupling $\Omega$ is switched on and applied until time $t=t_1$, at which point it is turned off, resulting in a portion of the population being transferred into mode $|b\rangle$. As the initial state was a coherent state in both modes, after the population transfer both $|a\rangle$ and $|b\rangle$ are described by coherent states, with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.

@@

After this hold time $\tau_{\mathrm{hold}} = t_2 - t_1 $, the coupling $\Omega$ is switched back on until time $t_3$, with a phase shift $\phi$ compared to its first application. During this last stage, the two modes exchange population and the quadrature variance is converted into number variance, in the same way that homodyne measurements are used in quantum optics to convert quadrature squeezing into number squeezing, which can be directly measured. In this interpretation, $\phi$ is the relative phase of the strong local oscillator, which allows specific phase angles of quadrature squeezing to be examined.  This experiment can also be interpreted as a Ramsey interferometer with a final beam splitter phase of $\phi$.

@@

This system can be solved analytically, and expressions for quadrature, absolute and relative number squeezing can be derived. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation}.
If we define
\begin{eqnarray}
s &=& \sin(\Omega (t_3 - t_2)) \label{eqsDef} \\
c &=& \cos(\Omega (t_3 - t_2)) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} / \hbar \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after applying the final coupling field are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \nonumber\\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) (n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B 
                     + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^*) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, e.g. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible. It should also be noted that, although there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences. This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

@@

In Figure \ref{figTwoModeAnalyticExamples} we plot a solution for the normalized number variance in mode $|a\rangle$ as a function of $\phi$, choosing a typical set of nonlinearities. This example illustrates the basic features of the system: During the final coupling period, the normalised number variance undergoes cyclic variation with a period $\tau=\pi / \Omega$, and has a series of minima. It is necessary to choose both the correct phase $\phi$ as well as the correct length of time for the coupling to be applied in order to obtain optimum squeezing.

This two-mode analytic model allows arbitrarily good number squeezing, provided there are no limits to the number of particles in the system or to the hold time. The two-mode approximation can be valid in tight traps with very low particle number, and this result correctly predicts the squeezing demonstrated by Gross \textit{et al.} \cite{Gross2010}. Unfortunately, for larger numbers of atoms in realistic traps there are more than two modes present, and multimode effects limit the level of squeezing that can be achieved.

@@

@@

An analytic solution to the multimode, higher-dimensional problem is intractable, so we must use a numerical solution.  We use stochastic methods based on phase space representations to simulate the dynamics of the multimode quantum fields.  These methods alleviate the problem of directly simulating elements of the Hilbert space, the size of which increases exponentially with the number of spatial modes \cite{stochasticRefs}.  Stochastic methods achieve this by finding a sufficiently well-behaved quasi-probability representation for the density matrix, which can then be simulated as the average behaviour of a number of low-dimensional samples.  These sample objects are the same size as the equivalent classical field.  In our case, the only tractable phase space representation is the functional Wigner representation, which is well behaved when the quantum field is approximately Gaussian, as it is for coherent and squeezed states.  Stochastic methods based on the functional Wigner representation have been used to model the behaviour of BECs and atom lasers \cite{johnssonET2007,dallET2009,dennisET2010}, as the Wigner representation allows the mapping of the system to a Fokker-Planck equation with positive semi-definite diffusion matrix (unlike the $P$-representation), and does not suffer exponential path-weighting problems that would constrain it to very short simulation intervals (as would the positive-$P$ representation).

The equation of motion for the density matrix is defined by the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}).  This defines the evolution of the functional Wigner distribution of the system, which has a one-to-one correspondence with the density matrix.  The equation of motion for the functional Wigner distribution contains derivatives of third order which we assume to be negligible.  This uncontrolled approximation is called the Truncated Wigner Approximation (TWA), and while it has been used widely on ultracold gases \cite{atomlaserWigner,orOtherUCGWigner}, care must be taken that the simulation remains valid. For our simulations we checked the results of our simulations against both the analytic two-mode solution, as well as checking for typical indications of TWA breakdown such as the appearance of negative densities in lightly populated modes.  We could also use the Bogoliubov analysis described in section \ref{sec:2D3Dsqueezing} as an independent estimate of the validity of the TWA.  We found that the TWA remained valid except in some 3D simulations with high nonlinearities, where there were a large number of modes, and the number of particles per mode could drop to ten or less in the densest regions where the squeezing was generated. Over very long time scales, this low mode occupation began to result in TWA breakdown.  Fortunately, the best squeezing was typically found in regions where the TWA was valid.

Under the TWA, the equation of motion for the functional Wigner distribution is a Fokker-Planck equation, and can therefore be sampled by a set of stochastic equations, as described in \cite{GardinerQuantumNoise}.  These stochastic fields $\phi(\mathbf{x})$ are related to field operator expectation values by  
\begin{equation}
\langle :\hat{\psi}^\dagger(\mathbf{x_1})\cdots\hat{\psi}^\dagger(\mathbf{x_n})\hat{\psi}(\mathbf{y_1})\cdots\hat{\psi}(\mathbf{y_m}):_{\mbox{sym}}\rangle = \mathbb{E}\left[\phi^*(\mathbf{x_1})\cdots\phi^*(\mathbf{x_n})\phi(\mathbf{y_1})\cdots\phi(\mathbf{y_m})\right]
\label{eqExpectationRelations}
\end{equation}
%
where $:\star:_{\mbox{sym}}$ denotes symmetric ordering of the operators, and $\mathbb{E}$ denotes a stochastic average over the variables $\phi(\mathbf{x})$.  While we see below that the evolution of these fields is deterministic, the initial state still requires a random element, so multiple realisations are required.  When approximating these equations on a discrete grid, the magnitude of initial noise, the relationship between stochastic averages and the expectation values, and the equations of motion all become grid-dependent.  This is not a signature of a fundamental problem, as the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}) assumes a contact potential between the atoms, which is not correct below a length scale that can probe the details of the true potentials.  Furthermore, it is not a problem in practice, as the calculation gives grid-independent predictions for physical observables well before that regime is reached.

@@

We integrate equations (\ref{eqStochasticEquations}) numerically in one, two and three spatial dimensions using the numerical package XMDS2 \cite{dennis2013}.  For the one- and two-dimensional simulations, the dimensional reduction is achieved by estimating the mode shape in three dimensions, and using dimensionally reduced values for $U_{ij}$ that match the zero-dimensional reduction given by Eq.~(\ref{eqChiUequivalence}).  To make relevant comparisons between equivalent situations with different numbers of spatial dimensions, this mode shape is typically chosen to match the chemical potentials of the initial states.

@@

\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/modal_squeezing_effects_1D.eps}
    \caption{Effects of mode shape on squeezing. The plot shows the normalized number variance in mode $|a \rangle$ as a function of the time the final coupling pulse is applied. We compare the analytic two-mode solution (thick red), the constant density mode (solid blue), the Thomas-Fermi mode (dashed) and the Gaussian mode (dash-dotted). The nonlinearities of the various 1D simulations have been adjusted for their mode shape so that they are equivalent to that of the 0D, two-mode mode nonlinearity. Parameters used were $n_a = n_b =1 \times 10^5$, $\chi_{aa}=0.03 \hbar, \chi_{ab}=\chi_{bb}=0, \tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=4.0$. Maximum squeezing is achieved when the atomic density profile is the most uniform.}
    \label{figModalSqueezingEffects1D}
\end{figure}

One way to avoid any requirement to engineer and maintain specific atomic modes is to post-select a spatial mode.  Rather than considering the variance in the total number of atoms in a mode, it may be possible to consider only the statistics of a spatially filtered area of the atomic cloud.  For example, suppose the atoms were in a Gaussian atomic distribution, but we consider a mode that only includes the atoms within $p$ standard deviations of the centre. For low $p$, this mode shape is considerably closer to the constant density case than a standard Gaussian. The equivalence between the single mode case and this mode is given by
\begin{equation}
\chi = U \frac{{\mathrm{Erf}} (\sqrt{2p})^3}{{\mathrm{Erf}}(p)^6} \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}}
\end{equation}
Spatial filtering can easily produce higher squeezing, but there is the obvious disadvantage that it  significantly reduces the effective number of atoms.  This is a serious issue for most applications of squeezed sources, such as interferometry, where the signal to noise scales with the square root of the flux.  We will focus on methods for producing squeezing with large atomic number.

\section{Using a $\pi$-pulse to improve mode-matching}
Modal mismatch is generated by the different atomic potentials seen by the two internal states.  A straightforward method of minimising this effect is to use a $\pi$-pulse to swap the populations of the internal states after half of the evolution. As can be seen by Eqs~(\ref{eqa1}) and (\ref{eqb1}), such a pulse also leaves the quantum statistics of each state unchanged. In the limit where each state remains in a single mode (although that mode might change), and exactly half of the atoms are in each state, this technique ensures that both modes have identical mode shapes after the hold time. In practice this will never work perfectly as the mode shapes will change, resulting in non-identical environments for the two modes. Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement. More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.

These points are illustrated in Fig.~\ref{piPulseFig}, which shows the best squeezing obtainable from a specific system with and without the $\pi$-pulse, as well as the extra squeezing that can be obtained by allowing a longer hold time as well as the $\pi$-pulse.
\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/pi_pulse.eps}
    \caption{Effect of a $\pi$-pulse halfway through the hold time. Common parameters used were $n_a = n_b =2 \times 10^5$, $\chi_{aa}=0.03\hbar$, $\chi_{ab}=\chi_{bb}=0$. Solid red line: Analytic two-mode solution, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.0$ (10.9dB squeezing); Solid blue line: no $\pi$-pulse, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.0$ (4.9dB squeezing);  Dashed blue line: with $\pi$-pulse, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.05$ (5.8dB squeezing); Dash-dotted blue line: with $\pi$-pulse and the longer hold time $\tau_{\mathrm{hold}}=5\times 10^{-4}$s allowed by the application of the pulse, $\phi=3.87$ (7.2dB squeezing).}
    \label{piPulseFig}
\end{figure}

\section{Squeezing in 2D and 3D} \label{sec:2D3Dsqueezing}

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=7cm]{figures/dimensional_effects_on_squeezing_1.eps}
    \label{figDimensionalSqueezingEffects:sub1}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=7cm]{figures/dimensional_effects_on_squeezing_2.eps}
    \label{figDimensionalSqueezingEffects:sub2}
  \end{subfigure}
\caption{Effects of dimension on squeezing in two different nonlinear regimes, assuming that the chemical potential is constant for all simulations. If two nonlinearities are non-zero, the deleterious effects of higher dimensions on squeezing are exacerbated. Thick red line, analytic two-mode solution; solid blue line, 1D; dashed line, 2D; dash-dotted line, 3D. (a) Parameters: $n_a = n_b =2 \times 10^5$, $\chi_{aa}=\chi_{bb}=0.03\hbar$, $\chi_{ab}=0$, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=6.2$. (b) $n_a = n_b =2 \times 10^5$, $\chi_{aa}=0.03\hbar$, $\chi_{bb}=\chi_{ab}=0$, $\tau_{\mathrm{hold}}=3\times 10^{-4}$s, $\phi=3.1$.}
  \label{figDimensionalSqueezingEffects}
\end{figure}

The simulations in the previous two sections were one dimensional, and show how having spatially constant density and using a $\pi$-pulse will improve the resultant squeezing.  While these conclusions remain valid for simulations in two or three dimensions, adding each additional dimension increases the number of spatial modes in a given energy range.  For BECs of different dimension at the same energy per particle, this leads to a degradation of squeezing, even when the potential (and the resulting atomic density) is constant across the bulk of the trap.  This is demonstrated in Fig.~\ref{figDimensionalSqueezingEffects}, where we show the squeezing generated in 1D, 2D and 3D BECs for two different parameter regimes. In both cases we take the BEC to be trapped in a box potential, which in the Thomas-Fermi limit leads to a constant density profile across the BEC. In Figure~\ref{figDimensionalSqueezingEffects}a we consider the case where $U_{aa}=U_{bb}\neq 0$, $U_{ab}=0$, while in Figure~\ref{figDimensionalSqueezingEffects}b we consider the case $U_{aa}\neq 0$, $U_{ab}=U_{bb}=0$. In both cases it is clear that number squeezing is best in lower dimensions, and becomes degraded or non-existent as we move to a three dimensional system; the effect is worse if more than one nonlinearity is non-zero.
